\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-03-01 jasa/paper-jasa.tex
Final draft

\subsection{Composite hypothesis testing}

@@

\subsection{Mixtures of 1D normals}

As a simple but illustrative example, let us first consider the approximation of
the likelihood log-ratio $\log \left( r(\mathbf{x};\theta_0,\theta_1) \right)$ between the mixtures
$p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$ illustrated in
Figure~\ref{fig:1} and defined as
\begin{align}
p(\mathbf{x}|\theta_0) &= (\frac{1}{2} - g) p_{c_0}(\mathbf{x}) + (\frac{1}{2} - g) p_{c_1}(\mathbf{x}) + g p_{c_2}(\mathbf{x}), \\
p(\mathbf{x}|\theta_1) &= \frac{1}{2} p_{c_0}(\mathbf{x}) + \frac{1}{2} p_{c_1}(\mathbf{x}),
\end{align}
where $p_{c_0} := {\cal N}(\mu=-2, \sigma^2=0.5625)$, $p_{c_1} := {\cal N}(\mu=0, \sigma^2=4)$,
$p_{c_2} := {\cal N}(\mu=1, \sigma^2=0.25)$ and where $g$ is a parameter of $p(\mathbf{x}|\theta_0)$
set to $0.05$. Samples drawn from $p(\mathbf{x}|\theta_0)$ are shown in the
figure and will be used later for inference.

@@

\subsection{Advanced example}

\glnote{todo}

@@

More generally, likelihood ratio testing directly relates to the density ratio
estimation problem, which consists in estimating the ratio of two probability
densities from given and finite collections of observations ${\cal D}_0$ and
${\cal D}_1$. Density ratio estimation is connected to many machine learning
fundamental problems, including transfer learning~\citep{sugiyama2012machine},
probabilistic classification and regression~\citep{vapnik1998statistical},
outlier detection~\citep{hido2011statistical}, and many others. For learning
under covariate shift, \cite{shimodaira2000improving} and \cite{sugiyama2005input} estimate
the density ratio $r(\mathbf{x};\theta_0,\theta_1) =
\frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)}$ from straightforward
approximations $\hat{p}(\mathbf{x}|\theta_0)$ and $\hat{p}(\mathbf{x}|\theta_1)$
separately obtained using kernel density estimation. Despite its theoretical
consistency, this approach is known to be ineffective in
practice~\citep{sugiyama2007covariate,bickel2009discriminative}, since
it relies on modeling numerator and denominator high-dimensional densities,
which is a harder problem than modeling their ratio only.
While the proposed method also proceeds in two similar steps, estimating
$p(s(\mathbf{x}))$ is much easier than estimating $p(\mathbf{x})$,
since $s$ projects $\mathbf{x}$ into a one-dimensional space in which only
the informative content of $r(\mathbf{x})$ is preserved.
Finally, in contrast with the
proposed method which decouples reduction from calibration, other
approaches proposed within the literature (see
\cite{sugiyama2012density,gretton2009covariate,nguyen2010estimating,vapnik2013constructive}
and references therein) provide solutions for estimating
$r(\mathbf{x};\theta_0,\theta_1)$ directly from $\mathbf{x}$, in one step. Under some
assumptions, the convergence of the obtained estimates is also proven for some
of these approaches.

@@

\section*{Acknowledgements} KC would like to thank Daniel Whiteson for
encouragement throughout the project and Alex Ihler for challenging discussions
that  led to a reformulation of the initial idea. KC would also like to thank
Shimon Whiteson, Babak Shahbaba for advice in the presentation, Radford Neal for
discussion of his earlier work, Yann LeCun, Philip Stark, and Pierre Baldi for
their feedback on the project early in its conception, Bal\'azs K\'egl for
feedback to the draft,  and Yuri
Shirman for reassuring cross checks of the Theorem.  KC is supported by the US
National Science Foundation grants PHY-0854724 and PHY-0955626.  KC is grateful
to UC-Irvine for their hospitality while this research was carried out and the
Moore and Sloan foundations for their generous support of the data science
environment at NYU.
\glnote{Update the acknowlegdments? What do we need to mention regarding my financial support?}

>>>>

\subsection{Composite hypothesis testing}
\label{sec:composite-hypothesis-testing}

@@

In this section, we illustrate the proposed method on two representative
examples and then discuss its application to high energy physics. For reproducibility and extended details, the source
code is available online\footnote{\url{https://github.com/diana-hep/carl}}.

\subsection{Likelihood ratios of mixtures of normals}

As a simple and illustrative example, let us first consider the approximation of
the likelihood log-ratio $\log \left( r(\mathbf{x};\theta_0,\theta_1) \right)$ between the 1D mixtures
$p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$ illustrated in
Figure~\ref{fig:1} and defined as
\begin{align}
p(\mathbf{x}|\theta_0) &= (\frac{1}{2} - g) p_{c_0}(\mathbf{x}) + (\frac{1}{2} - g) p_{c_1}(\mathbf{x}) + g p_{c_2}(\mathbf{x}), \\
p(\mathbf{x}|\theta_1) &= \frac{1}{2} p_{c_0}(\mathbf{x}) + \frac{1}{2} p_{c_1}(\mathbf{x}),
\end{align}
where $p_{c_0} := {\cal N}(\mu=-2, \sigma^2=0.5625)$, $p_{c_1} := {\cal N}(\mu=0, \sigma^2=4)$,
$p_{c_2} := {\cal N}(\mu=1, \sigma^2=0.25)$ and where $g$ is a parameter of $p(\mathbf{x}|\theta_0)$
set to $0.05$. Samples drawn from $p(\mathbf{x}|\theta_0)$ are shown in the
figure and will be used later for inference.

@@

\subsection{Parameterized inference from multidimensional data}

Let us now consider the more challenging problem of maximum likelihood
inference from multidimensional data generated by a black-box simulator.
For the sake of the illustration, we will assume 5-dimensional observed samples $\mathbf{x}$ generated
from the following process $p_0$:
    \begin{enumerate}
        \item $\mathbf{z} := (z_0, z_1, z_2, z_3, z_4)$, such that
            $z_0 \sim {\cal N}(\mu=A, \sigma=1)$,
            $z_1 \sim {\cal N}(\mu=B, \sigma=3)$,
            $z_2 \sim {\text{Mixture}}({\cal N}(\mu=-2, \sigma=1), {\cal N}(\mu=2, \sigma=0.5))$,
            $z_3 \sim {\text{Exponential}(\lambda=3)}$ and
            $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
        \item $\mathbf{x} := R  \mathbf{z}$, where $R$ is a fixed
        semi-positive definite $5 \times 5$ matrix defining a fixed projection
        of $\mathbf{z}$ into the observed space.
    \end{enumerate}
Given observations ${\cal D}$ represented in Figure~\ref{fig:4},
our goal is to infer the nominal values $A=1$ and $B=-1$ that
were used for generating this data.

\begin{figure}
    \centering
    \includegraphics[width=1.0\textwidth]{figures/fig4.pdf}
    \caption{Observed 5-dimensional data ${\cal D}$ for nominal values $A=1$ and $B=-1$.}
    \label{fig:4}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=1.0\textwidth]{figures/fig5.pdf}
    \caption{Parameter inference from parameterized approximated likelihood ratios.
             The constructed objective is minimized with Bayesian optimization.
             Red dots correspond to sampled candidate pairs $(A,B)$.
             The blue dot corresponds to the minimum found after $50$ iterations. Nominal values are $A=1$ and $B=-1$.}
    \label{fig:5}
\end{figure}

As previously, we make use of Eqn.~\ref{eq:mle_withs} for parameter inference
with approximated likelihood ratios and fix the alternate to $\theta_1:=(A=0,
B=0)$. Following recommendations from Section~\ref{sec:param-clf} to reduce
computational requirements, we train a single parameterized classifier
$s(\mathbf{x};\theta:=(A,B),\theta_1)$ (i.e., a neural network with 2 hidden layers of 40 nodes) and whose output is calibrated on-the-fly
with isotonic regression for every candidate parameter pair $\theta:=(A,B)$.
Minimizing the mean negative log of the constructed objective with Bayesian
optimization~\citep{brochu2010tutorial} eventually leads to the estimated
posterior mean and variance shown in Figure~\ref{fig:5}. The objective is minimized after $50$ iterations
at $(\hat A=1.00,\hat B=-1.15)$. Again, repeated experiments (omitted here) confirm that
the proposed procedure yields an unbiased maximum likelihood estimator.

@@

More generally, likelihood ratio testing directly relates to the density ratio
estimation problem, which consists in estimating the ratio of two
densities from finite collections of observations ${\cal D}_0$ and
${\cal D}_1$. Density ratio estimation is connected to many machine learning
fundamental problems, including transfer learning~\citep{sugiyama2012machine},
probabilistic classification and regression~\citep{vapnik1998statistical},
outlier detection~\citep{hido2011statistical}, and many others. For learning
under covariate shift, \cite{shimodaira2000improving} and \cite{sugiyama2005input} estimate
the density ratio $r(\mathbf{x};\theta_0,\theta_1) =
\frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)}$ from straightforward
approximations $\hat{p}(\mathbf{x}|\theta_0)$ and $\hat{p}(\mathbf{x}|\theta_1)$
separately obtained using kernel density estimation. Despite its theoretical
consistency, this approach is known to be ineffective in
practice~\citep{sugiyama2007covariate,bickel2009discriminative}, since
it relies on modeling numerator and denominator high-dimensional densities,
which is a harder problem than modeling their ratio only.
While the proposed method also proceeds in two similar steps, estimating
$p(s(\mathbf{x}))$ is much easier than estimating $p(\mathbf{x})$,
since $s$ projects $\mathbf{x}$ into a one-dimensional space in which only
the informative content of $r(\mathbf{x})$ is preserved.
Finally, in contrast with the
proposed method which decouples reduction from calibration, other
approaches proposed within the literature (see
\cite{sugiyama2012density,gretton2009covariate,nguyen2010estimating,vapnik2013constructive}
and references therein) provide solutions for estimating
$r(\mathbf{x};\theta_0,\theta_1)$ directly from $\mathbf{x}$, in one step. Under some
assumptions, the convergence of the obtained estimates is also proven for some
of these approaches.

@@

\section*{Acknowledgements} KC and GL are both supported through NSF ACI-1450310.
KC would like to thank Daniel Whiteson for
encouragement and Alex Ihler for challenging discussions
that  led to a reformulation of the initial idea. KC would also like to thank
Shimon Whiteson, Babak Shahbaba for advice in the presentation, Radford Neal for
discussion of his earlier work, Yann LeCun, Philip Stark, and Pierre Baldi for
their feedback on the project early in its conception, Bal\'azs K\'egl for
feedback to the draft,  and Yuri
Shirman for reassuring cross checks of the Theorem.   KC is grateful
to UC-Irvine for their hospitality while this research was carried out and the
Moore and Sloan foundations for their generous support of the data science
environment at NYU.
